\vspace{30mm} {\Large\bf RISC-V Duo-PLIC for the Advancced Interrupt Architecture} \\ {\large Document Version \DuoPLICRev}

### 2023-06-16 doc/src/CSRs.tex

CSRs \z{miselect} and \z{mireg} provide a window for accessing multiple
registers beyond the CSRs in Table~\ref{tab:CSRs-M}.
The value of \z{miselect} determines which register is currently
accessible through alias CSR \z{mireg}.
\z{miselect} is a {\WARL} register, and it must support a minimum range
of values depending on the implemented features.
When an IMSIC is not implemented, \z{miselect} must be able to hold at
least any \mbox{6-bit} value in the range 0 to \z{0x3F}.
When an IMSIC is implemented, \z{miselect} must be able to hold any
\mbox{8-bit} value in the range 0 to \z{0xFF}.
Values for \z{miselect} in the range 0 to \z{0xFF} are currently
assigned in subranges as follows:
\begin{displayLinesTable}[l@{\quad}l]
\z{0x00}--\z{0x2F} & reserved \\
\z{0x30}--\z{0x3F} & major interrupt priorities \\
\z{0x40}--\z{0x6F} & reserved \\
\z{0x70}--\z{0xFF} & external interrupts (only with an IMSIC) \\
\end{displayLinesTable}
\z{miselect} may also support values outside the range
\z{0x00}--\z{0xFF}, though no standard registers are currently
allocated to values above \z{0xFF}.

@@

The space of registers accessible through the \z{siselect}/\z{sireg}
window is separate from but parallels that of machine level, being for
supervisor-level interrupts instead of machine-level interrupts.
The allocated values for \z{siselect} in the range 0 to \z{0xFF} are
once again these:
\begin{displayLinesTable}[l@{\quad}l]
\z{0x00}--\z{0x2F} & reserved \\
\z{0x30}--\z{0x3F} & major interrupt priorities \\
\z{0x40}--\z{0x6F} & reserved \\
\z{0x70}--\z{0xFF} & external interrupts (only with an IMSIC) \\
\end{displayLinesTable}
For maximum compatibility, it is recommended that \z{siselect} support
at least a \mbox{9-bit} range, 0 to \z{0x1FF}, regardless of whether an
IMSIC exists.

@@

The new VS CSRs in the table (\z{vsiselect}, \z{vsireg},
\z{vstopei}, and \z{vstopi})
all match supervisor CSRs, and substitute for those
supervisor CSRs when executing in a virtual machine (in \mbox{VS-mode}
or \mbox{VU-mode}).

@@

The space of registers selectable by \z{vsiselect} is more limited than
for machine and supervisor levels:
\begin{displayLinesTable}[l@{\quad}l]
\z{0x000}--\z{0x02F} & reserved \\
\z{0x030}--\z{0x03F} & inaccessible \\
\z{0x040}--\z{0x06F} & reserved \\
\z{0x070}--\z{0x0FF} & external interrupts (IMSIC only), or inaccessible \\
\z{0x100}--\z{0x1FF} & reserved \\
\end{displayLinesTable}

For alias CSRs \z{sireg} and \z{vsireg}, the hypervisor extension's
usual rules for when to raise a virtual instruction exception (based
on whether an instruction is \emph{HS-qualified}) are not applicable.

The rules for \z{sireg} and \z{vsireg} in this section assume
\z{vsiselect} is accessible from \mbox{HS-mode} as usual.
See Section~\ref{sec:CSRs-stateen} for when
\z{vsiselect} may not be accessible from \mbox{HS-mode}.

All attempts to access \z{vsireg} from \mbox{VS-mode}
or \mbox{VU-mode} raise a virtual instruction exception.

When \z{vsiselect} has a reserved value (including values above
\z{0x1FF} not designated for custom use), attempts from \mbox{M-mode}
or \mbox{HS-mode} to access \z{vsireg}, or from \mbox{VS-mode}
or \mbox{VU-mode} to access \z{sireg} (really \z{vsireg}),
should preferably raise an illegal instruction exception.

When \z{vsiselect} has the number of an \emph{inaccessible} register,
attempts from \mbox{M-mode} or \mbox{HS-mode} to access \z{vsireg}
raise an illegal instruction exception, and attempts from
\mbox{VS-mode} or \mbox{VU-mode} to access \z{sireg}
(really \z{vsireg}) raise a virtual instruction exception.

@@

The locations of registers for external interrupts (numbers
\z{0x70}--\z{0xFF}) are accessible only when field VGEIN of \z{hstatus}
is the number of an implemented guest external interrupt, not zero.
If VGEIN is not the number of an implemented guest external interrupt
(including the case when no IMSIC is implemented), then all
indirect register numbers in the ranges \z{0x030}--\z{0x03F} and
\z{0x070}--\z{0x0FF} designate an inaccessible register at VS level.

@@

The new hypervisor CSRs in Table~\ref{tab:CSRs-hypervisor}
(\z{hvien}, \z{hvictl}, \z{hviprio1}, and \z{hviprio2})
augment \z{hvip} for injecting interrupts into VS level.
The use of these registers is covered in Chapter~\ref{ch:VSLevel} on
interrupts for virtual machines.

If extension Sscsrind is also implemented, then when \z{vsiselect} has
a value in the range \z{0x30}--\z{0x3F} or \z{0x70}--\z{0xFF},
attempts from privilege modes other than VS or VU to access alias CSRs
\z{vsireg2} through \z{vsireg6} raise an illegal instruction exception,
and attempts from VS or \mbox{VU-mode} to access \z{sireg2}
through \z{sireg6} or \z{vsireg2} through \z{vsireg6}
raise a virtual instruction exception.

@@

@@

If extension Smstateen is implemented together with the
Advanced Interrupt Architecture (AIA), three bits of state-enable
register \z{mstateen0} control access to AIA-added state
from privilege modes less privileged than \mbox{M-mode}:
\begin{displayLinesTable}[l@{\quad}l]
bit 60 & CSRs \z{siselect}, \z{sireg}, \z{vsiselect}, and \z{vsireg} \\
bit 59 & all other state added by the AIA
          and not controlled by bits 60 and 58 \\
bit 58 & all IMSIC state, including CSRs \z{stopei} and \z{vstopei} \\
\end{displayLinesTable}

@@

If the hypervisor extension is implemented, the same three bits are
defined also in hypservisor CSR \z{hstateen0}
but concern only the state potentially
accessible to a virtual machine executing in privilege modes VS and VU:
\begin{displayLinesTable}[l@{\quad}l]
bit 60 & CSRs \z{siselect} and \z{sireg}
          (really \z{vsiselect} and \z{vsireg}) \\
bit 59 & CSRs \z{siph} and \z{sieh} (RV32 only) and \z{stopi}
          (really \z{vsiph}, \z{vsieh}, and \z{vstopi}) \\
bit 58 & all state of IMSIC guest interrupt files,
          including CSR \z{stopei} (really \z{vstopei}) \\
\end{displayLinesTable}

If one of these bits is zero in \z{hstateen0},
and the same bit is one in \z{mstateen0},
then an attempt to access the corresponding state
from VS or \mbox{VU-mode} raises a virtual instruction exception,
unless one of the following cases applies
to cause an illegal instruction exception instead:
\begin{itemize}

\item
An attempt to access \z{sireg} raises an illegal instruction exception
if the same attempt would do so when bit~60 of \z{hstateen0} is one.

\item
When $\mbox{XLEN} > \mbox{32}$, attempts to access high-half CSRs
\z{siph} and \z{sieh} always raise an illegal instruction exception.

\end{itemize}

>>>>

CSRs \z{miselect} and \z{mireg} provide a window for accessing multiple
registers beyond the CSRs in Table~\ref{tab:CSRs-M}.
The value of \z{miselect} determines which register is currently
accessible through alias CSR \z{mireg}.
\z{miselect} is a {\WARL} register, and it must support a minimum range
of values depending on the implemented features.
When an IMSIC is not implemented, \z{miselect} must be able to hold at
least any \mbox{6-bit} value in the range 0 to \z{0x3F}.
When an IMSIC is implemented, \z{miselect} must be able to hold any
\mbox{8-bit} value in the range 0 to \z{0xFF}.
Values for \z{miselect} in the range 0 to \z{0xFF} are currently
assigned in subranges as follows:\nopagebreak
\begin{displayLinesTable}[l@{\quad}l]
\z{0x00}--\z{0x2F} & reserved \\
\z{0x30}--\z{0x3F} & major interrupt priorities \\
\z{0x40}--\z{0x6F} & reserved \\
\z{0x70}--\z{0xFF} & external interrupts (only with an IMSIC) \\
\end{displayLinesTable}
\z{miselect} may also support values outside the range
\z{0x00}--\z{0xFF}, though no standard registers are currently
allocated to values above \z{0xFF}.

@@

The space of registers accessible through the \z{siselect}/\z{sireg}
window is separate from but parallels that of machine level, being for
supervisor-level interrupts instead of machine-level interrupts.
The allocated values for \z{siselect} in the range 0 to \z{0xFF} are
once again these:\nopagebreak
\begin{displayLinesTable}[l@{\quad}l]
\z{0x00}--\z{0x2F} & reserved \\
\z{0x30}--\z{0x3F} & major interrupt priorities \\
\z{0x40}--\z{0x6F} & reserved \\
\z{0x70}--\z{0xFF} & external interrupts (only with an IMSIC) \\
\end{displayLinesTable}
For maximum compatibility, it is recommended that \z{siselect} support
at least a \mbox{9-bit} range, 0 to \z{0x1FF}, regardless of whether an
IMSIC exists.

@@

The new hypervisor CSRs in the table (\z{hvien},
\z{hvictl}, \z{hviprio1}, and \z{hviprio2}) augment
\z{hvip} for injecting interrupts into VS level.
The use of these registers is covered in Chapter~\ref{ch:VSLevel} on
interrupts for virtual machines.

The new VS CSRs (\z{vsiselect}, \z{vsireg},
\z{vstopei}, and \z{vstopi})
all match supervisor CSRs, and substitute for those
supervisor CSRs when executing in a virtual machine (in \mbox{VS-mode}
or \mbox{VU-mode}).

@@

The space of registers selectable by \z{vsiselect} is more limited than
for machine and supervisor levels:\nopagebreak
\begin{displayLinesTable}[l@{\quad}l]
\z{0x000}--\z{0x02F} & reserved \\
\z{0x030}--\z{0x03F} & inaccessible \\
\z{0x040}--\z{0x06F} & reserved \\
\z{0x070}--\z{0x0FF} & external interrupts (IMSIC only), or inaccessible \\
\z{0x100}--\z{0x1FF} & reserved \\
\end{displayLinesTable}

For alias CSRs \z{sireg} and \z{vsireg}, the hypervisor extension's
usual rules for when to raise a virtual instruction exception (based
on whether an instruction is \emph{HS-qualified}) are not applicable.
The rules given in this section for \z{sireg} and \z{vsireg}
apply instead, unless overridden by the requirements of
Section~\ref{sec:CSRs-stateen}, which take precedence over
this section when extension Smstateen is also implemented.

A virtual instruction exception is raised for attempts from
\mbox{VS-mode} or \mbox{VU-mode} to directly access \z{vsireg},
or attempts from \mbox{VU-mode} to access \z{sireg}.

When \z{vsiselect} has a reserved value (including values above
\z{0x1FF} not designated for custom use), attempts from \mbox{M-mode}
or \mbox{HS-mode} to access \z{vsireg}, or from \mbox{VS-mode}
to access \z{sireg} (really \z{vsireg}),
should preferably raise an illegal instruction exception.

When \z{vsiselect} has the number of an \emph{inaccessible} register,
attempts from \mbox{M-mode} or \mbox{HS-mode} to access \z{vsireg}
raise an illegal instruction exception, and attempts from
\mbox{VS-mode} to access \z{sireg}
(really \z{vsireg}) raise a virtual instruction exception.

@@

The indirectly accessed registers for external interrupts (numbers
\z{0x70}--\z{0xFF}) are accessible only when field VGEIN of \z{hstatus}
is the number of an implemented guest external interrupt, not zero.
If VGEIN is not the number of an implemented guest external interrupt
(including the case when no IMSIC is implemented), then all
indirect register numbers in the ranges \z{0x030}--\z{0x03F} and
\z{0x070}--\z{0x0FF} designate an inaccessible register at VS level.

@@

If extension Sscsrind is also implemented, then when \z{vsiselect} has
a value in the range \z{0x30}--\z{0x3F} or \z{0x70}--\z{0xFF},
attempts from \mbox{M-mode} or \mbox{HS-mode} to access alias CSRs
\z{vsireg2} through \z{vsireg6} raise an illegal instruction exception,
and attempts from \mbox{VS-mode} to access \z{sireg2} through \z{sireg6}
raise a virtual instruction exception.

@@

Instructions that read/write CSR \z{stopei} or \z{vstopei} are
considered to be \emph{HS-qualified} unless all of following are true:
the hart has an \mbox{IMSIC}, extension Smstateen is
implemented, and bit~58 of \z{mstateen0} is zero.
(See the next section, \ref{sec:CSRs-stateen}, about \z{mstateen0}.)

@@

If extension Smstateen is implemented together with the
Advanced Interrupt Architecture (AIA), three bits of state-enable
register \z{mstateen0} control access to AIA-added state
from privilege modes less privileged than \mbox{M-mode}:\nopagebreak
\begin{displayLinesTable}[l@{\quad}l]
bit 60 & CSRs \z{siselect}, \z{sireg}, \z{vsiselect}, and \z{vsireg} \\
bit 59 & all other state added by the AIA
          and not controlled by bits 60 and 58 \\
bit 58 & all IMSIC state, including CSRs \z{stopei} and \z{vstopei} \\
\end{displayLinesTable}

@@

If the hart does not have an \mbox{IMSIC}, bit~58 of \z{mstateen0}
is read-only zero, but Smstateen has no effect on
attempts to access the nonexistent \mbox{IMSIC} state.

\begin{commentary}
This means in particular that, when the hart does not have an
\mbox{IMSIC}, the following raise a virtual instruction exception
as described in Section~\ref{ch:CSRs-hypervisor}, not an illegal
instruction exception, despite that bit~58 of\/ \z{mstateen0} is zero:
\begin{tightList}
\item
attempts from \mbox{VS-mode} to access\/ \z{sireg}
(really\/ \z{vsireg}) while\/ \z{vsiselect} has
a value in the range\/ \z{0x70}--\z{0xFF}; and
\item
attempts from \mbox{VS-mode} to access\/
\z{stopei} (really\/ \z{vstopei}).
\end{tightList}
\end{commentary}

If bit~60 of \z{mstateen0} is one, then regardless of any other
\z{mstateen} bits (including bits 58 and 59 of \z{mstateen0}),
a virtual instruction exception is raised as described in
Section~\ref{ch:CSRs-hypervisor} for all attempts from
\mbox{VS-mode} or \mbox{VU-mode} to directly access \z{vsireg},
and for all attempts from \mbox{VU-mode} to access \z{sireg}.
This behavior is overridden only when bit~60 of \z{mstateen0} is zero.

If the hypervisor extension is implemented, the same three bits are
defined also in hypervisor CSR \z{hstateen0}
but concern only the state potentially
accessible to a virtual machine executing
in privilege modes VS and VU:\nopagebreak
\begin{displayLinesTable}[l@{\quad}l]
bit 60 & CSRs \z{siselect} and \z{sireg}
          (really \z{vsiselect} and \z{vsireg}) \\
bit 59 & CSRs \z{siph} and \z{sieh} (RV32 only) and \z{stopi}
          (really \z{vsiph}, \z{vsieh}, and \z{vstopi}) \\
bit 58 & all state of IMSIC guest interrupt files,
          including CSR \z{stopei} (really \z{vstopei}) \\
\end{displayLinesTable}

If one of these bits is zero in \z{hstateen0},
and the same bit is one in \z{mstateen0},
then an attempt to access the corresponding state
from VS or \mbox{VU-mode} raises a virtual instruction exception.
(But note that, for high-half CSRs \z{siph} and
\z{sieh}, this applies only when XLEN =~32.
When $\mbox{XLEN} > \mbox{32}$, an attempt to access
\z{siph} or \z{sieh} raises an illegal instruction
exception as usual, not a virtual instruction exception.)

If bit~60 is one in \z{mstateen0} but is zero in \z{hstateen0},
then all attempts from VS or \mbox{VU-mode} to access
\z{siselect} or \z{sireg} raise a virtual instruction
exception, not an illegal instruction exception, regardless
of the value of \z{vsiselect} or any other \z{mstateen} bits.
